---
title: "EGFR_mutant"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}

# load library
library("limma")
library("edgeR")
library("glmnet")
library("factoextra")
library("FactoMineR")
library("caret")
library("SummarizedExperiment")
library("gplots")
library("survival")
library("survminer")
library("RColorBrewer")
library("gProfileR")
library("genefilter")
library('gprofiler2')
library('dplyr')
library(pheatmap)
library(dendextend)

# Clear the entire environment
#rm(list = ls())

```

### Extract tumor patients with EGFR mutation

```{r, warning=FALSE, message=FALSE}
tcga_data = readRDS(file = "data/tcga_luad_data.RDS")
tcga_data_disease = tcga_data[,colData(tcga_data)$definition == "Primary solid Tumor"] # KEEP TUMOR
```


```{r, warning=FALSE, message=FALSE}

# import mutant info
EGFR_patient <- read.table("data/explore-case-table-2023-12-06.tsv",skip=1, sep = "\t")
colnames(EGFR_patient)[1] <- "patient"

# KEEP  patients with EGFR mutation
tcga_data_disease$is_EGFR_mutant = tcga_data_disease$patient %in% EGFR_patient$patient
tcga_EGFR = tcga_data_disease[, tcga_data_disease$is_EGFR_mutant]

# remove outliers based on initial NMF
tcga_EGFR = tcga_EGFR[, !tcga_EGFR$barcode%in% c("TCGA-44-5645-01B-04R-A277-07", "TCGA-44-6147-01B-06R-A277-07")]
## above two patients looks like outlier based on initial NMF

```

### Remove patients with obscure pattern in gene expression
```{r, warning=FALSE, message=FALSE}

  ## pure data normalization using limma
  dge = DGEList(counts=assay(tcga_EGFR),
                samples=colData(tcga_EGFR),
                genes=as.data.frame(rowData(tcga_EGFR)))
  keep = filterByExpr(dge, design = NULL)# defining which genes to keep
  dge = dge[keep,,keep.lib.sizes=FALSE]# filtering the dge object
  rm(keep)# 
  # Normalization (TMM followed by voom)
  dge = calcNormFactors(dge, method="TMM") #Trimmed Mean of M-values - > remove batch effects
  v = voom(dge, design = NULL, plot=FALSE) #mean-variance modelling at the observational level
  
  ### apply NMF to examine clusters
  log2.norm.count = log2(2^v$E + 1)
  library(NMF)
  res = nmf(log2.norm.count, rank = 2)
  V.hat <- fitted(res)
  w <- basis(res)
  h <- coef(res)

  
  my_hclust_subj <- hclust(dist(t(h)), method = "complete")
  my_subj_cluster <- cutree(tree = as.dendrogram(my_hclust_subj), k = 5)
  my_subj_cluster <- data.frame(cluster = 
                                ifelse(test = my_subj_cluster == 4, 
                                       yes = "cluster 1", 
                                       no = "cluster 2"))
pheatmap(h,
         annotation_col = my_subj_cluster,
         fontsize_row = 7,
         show_colnames = FALSE)

my_subj_cluster <- cutree(tree = as.dendrogram(my_hclust_subj), k = 5)
rm.ID = names(which(my_subj_cluster == 4)) ### remove those patients that cannot be clearly clustered.

###
tcga_EGFR = tcga_EGFR[, !tcga_EGFR$barcode%in%rm.ID] 
###

```


### double check whether the unclear patients are cleaned 
```{r, warning=FALSE, message=FALSE}

   #tcga_EGFR = readRDS(file = "tcga_EGFR_filtered.RDS")
   dge = DGEList(counts=assay(tcga_EGFR),
                samples=colData(tcga_EGFR),
                genes=as.data.frame(rowData(tcga_EGFR)))
  
  # filtering
  keep = filterByExpr(dge,design = NULL)# defining which genes to keep
  dge = dge[keep,,keep.lib.sizes=FALSE]# filtering the dge object
  rm(keep)# 
  dge = calcNormFactors(dge,method="TMM") 
  v = voom(dge, design = NULL, plot=FALSE) 
  log2.norm.count = log2(2^v$E + 1)
  
  ### apply NMF
  library(NMF)
  res = nmf(log2.norm.count, rank = 2)
  V.hat <- fitted(res)
  w <- basis(res)
  h <- coef(res)
 # pheatmap(h, cluster_cols = TRUE)
  my_hclust_subj <- hclust(dist(t(h)), method = "complete")
  my_subj_cluster <- cutree(tree = as.dendrogram(my_hclust_subj), k = 2)
  my_subj_cluster <- data.frame(cluster = 
                                ifelse(test = my_subj_cluster == 1, 
                                       yes = "cluster 1", 
                                       no = "cluster 2"))
pheatmap(h,
         annotation_col = my_subj_cluster,
        fontsize_row = 7,
        show_colnames = FALSE)

### then create label for each patient based on clustering
  my_hclust_subj <- hclust(dist(t(h)), method = "complete")
  my_subj_cluster <- cutree(tree = as.dendrogram(my_hclust_subj), k = 2)
  
  all(names(my_subj_cluster) == tcga_EGFR$barcode)
  tcga_EGFR$Cluster = my_subj_cluster
  saveRDS(tcga_EGFR, file = "data/tcga_EGFR_filtered.RDS")
  
```

### PCA on cleaned data
```{r, warning=FALSE, message=FALSE}

  group = factor(my_subj_cluster)
  pca = prcomp(t(v$E))
  plot(pca$x[,1:2], pch=19, col = group)
  legend("bottomleft", inset=.01, legend = paste0("Cluster ", levels(group)), 
         pch=19, col=1:length(levels(group)))
  ## save
  pdf(file = "output/EGFRmutant_PCA_TwoCluster.pdf", width = 5, height = 5)
  plot(pca$x[,1:2], pch=19, col = group)
  legend("bottomleft",  inset=.01, legend = paste0("Cluster ", levels(group)),
         pch=19, col=1:length(levels(group)))
  dev.off()
  
```

### Call DE gene using limma, based on preprocessed data
```{r, warning=FALSE, message=FALSE}

source("limma_pipeline.R")
#return normalized data of mutants, model and top 100 genes(EGFR mutant)
limma_res_EGFR = limma_pipeline(
  tcga_data=tcga_EGFR,
  condition_variable="Cluster",
  reference_group="1",
  topNumber = 1000)
limma_res_EGFR$Cluster = tcga_EGFR$Cluster
saveRDS(object = limma_res_EGFR,
        file = "limma_res_EGFR.RDS",
        compress = FALSE)
```


#### Take a look at DE genes between the two Clusters
```{r, warning=FALSE, message=FALSE}

limma_res_EGFR = readRDS(file = "limma_res_EGFR.RDS")
topGenes = limma_res_EGFR$topGenes
topGenes = topGenes[,c("gene_name", "logFC", "AveExpr","t" , "adj.P.Val")]
topGenes$HighInCluster2 = ifelse(topGenes$logFC > 0, TRUE, FALSE)

library(dplyr)
limma_topGene <- topGenes %>% arrange(HighInCluster2, adj.P.Val)
sum(limma_topGene$logFC < 0)
limma_topGene[c(1:25, 517:541),]

```

#### save counts of DE top genes, both raw and normalized, into a table
```{r}
## column ordered by cluster: 
### the first 1:29 columns are cluster 1, and the remaining 27 columns are cluster 2
all(colnames(limma_res_EGFR$voomObj$E) == colnames( assay(tcga_EGFR)))
rowcounts = assay(tcga_EGFR)[rownames(limma_topGene), order(tcga_EGFR$Cluster)]
Normcounts = round(2^limma_res_EGFR$voomObj$E[rownames(limma_topGene), order(tcga_EGFR$Cluster)],2)
all(rownames(rowcounts) == rownames(limma_topGene))
all(rownames(Normcounts) == rownames(limma_topGene))

rownames(rowcounts) = rownames(Normcounts) = limma_topGene$gene_name
write.csv(rowcounts, file = "output/EGFR_Top1000_DEgenes_rowCount.csv")
write.csv(Normcounts, file = "output/EGFR_Top1000_DEgenes_NormCount.csv")

```



#### visualize the pattern, and annotate both rows and columns
```{r, warning=FALSE, message=FALSE}
# top genes selected by limma 
library(pheatmap)
##### reorder E
Clust1.ID = which(limma_res_EGFR$Cluster == 1)
Clust2.ID = which(limma_res_EGFR$Cluster == 2)
log2.norm.count = limma_res_EGFR$voomObj$E
colnames(log2.norm.count)[Clust1.ID] = paste0("C1_", Clust1.ID)
colnames(log2.norm.count)[Clust2.ID] = paste0("C2_", Clust2.ID)

ord.log2.norm.count = log2.norm.count[,order(colnames(log2.norm.count))]

use.geneID = c(1:25, 517:541)
use.count = ord.log2.norm.count[rownames(limma_topGene)[use.geneID],]
rownames(use.count) = limma_topGene$gene_name[use.geneID]

cal_z_score <- function(x){ (x - mean(x)) / sd(x)}
use.count.zscore <- t(apply(use.count, 1, cal_z_score))
#pheatmap(use.count.zscore, cluster_cols = FALSE)
#pheatmap(use.count.zscore)


library(dendextend)
my_hclust_gene <- hclust(dist(use.count.zscore), method = "complete")
my_gene_col <- cutree(tree = as.dendrogram(my_hclust_gene), k = 2)
my_gene_col <- data.frame(Genes = ifelse(test = my_gene_col == 1, yes = "cluster 1", no = "cluster 2"))

my_sample_col <- data.frame(Patients = rep(c("Cluster 1", "Cluster 2"), c(length(Clust1.ID),length(Clust2.ID))))
row.names(my_sample_col) <- colnames(ord.log2.norm.count)
pheatmap(use.count.zscore, 
        annotation_row = my_gene_col, 
         annotation_col = my_sample_col, 
        show_colnames = FALSE,
        fontsize_row = 5)

my_hclust_subj <- hclust(dist(t(use.count.zscore)), method = "complete")
my_subj_cluster <- cutree(tree = as.dendrogram(my_hclust_subj), k = 2)
table(my_subj_cluster, my_sample_col$Patients)### perfect separation

### save the picture
# pdf(file = "EGFRmutant_Heatmap_TwoCluster.pdf", height = 7, width = 7)
# print(pheatmap(use.count.zscore, 
#          annotation_row = my_gene_col, 
#          annotation_col = my_sample_col,
#         fontsize_row = 7,
#         show_colnames = FALSE))
# dev.off()

heatmap_filename <- "EGFRmutant_Heatmap_TwoCluster.png"
save_heatmap <- function(filename) {
  pheatmap(use.count.zscore,
         annotation_row = my_gene_col,
         annotation_col = my_sample_col,
        fontsize_row = 7,
        show_colnames = FALSE,
        filename = heatmap_filename)

}
save_heatmap(heatmap_filename)
#knitr::include_graphics(heatmap_filename)
```

### Significant difference in survival time between the two clusters
```{r, warning=FALSE, message=FALSE}

clinical = tcga_EGFR@colData
# we are only interested in the "Primary solid Tumor" cases for survival
clin_df = clinical[, c("patient","vital_status", 
                       "days_to_death", "days_to_last_follow_up",
                       "ajcc_pathologic_stage", "gender", "Cluster")]
dim(clin_df)

# create a new boolean variable that has TRUE for dead patients
# and FALSE for live patients
clin_df$deceased = clin_df$vital_status == "Dead"
clin_df$overall_survival = ifelse(clin_df$deceased,
                                  clin_df$days_to_death,
                                  clin_df$days_to_last_follow_up)

clin_df$Cluster1 <- ifelse(clin_df$Cluster == 1, TRUE, FALSE)

# fit a survival model
fit = survfit(Surv(overall_survival, deceased) ~ Cluster1, data=clin_df)
print(fit)
ggsurvplot(fit, data=clin_df,pval=T)

```


### Check whether cluster label is associated with gender
```{r, message=FALSE, warning=FALSE}

confusion.mat = table(clinical$gender, clinical$Cluster)
#OR >1 
dat = data.frame(y = ifelse(clinical$Cluster == "1", 0, 1),
                 x = clinical$gender)
res = glm(y~x, data = dat, family = "binomial")
# association is not significant
summary(res)

```

### check whether cluster label is associated with age
```{r, message=FALSE, warning=FALSE}

dat = data.frame(y = ifelse(clinical$Cluster == "1", 0, 1),
                 x = clinical$age_at_diagnosis)
res = glm(y~x, data = dat, family = "binomial")
summary(res)
# association is not significant
boxplot(clinical$age_at_diagnosis ~ clinical$Cluster, 
        xlab = "Cluster", ylab = "age at diagnosis")

```


### check whether cluster label is associated with weight
```{r, message=FALSE, warning=FALSE}

dat = data.frame(y = ifelse(clinical$Cluster == "1", 0, 1),
                 x = clinical$initial_weight)
res = glm(y~x, data = dat, family = "binomial")
summary(res)
boxplot(clinical$initial_weight ~ clinical$Cluster, 
        xlab = "Cluster", ylab = "Initial weight")
### not meaningful because a lot of missing value

```


### check whether cluster label is associated with $ajcc\_pathologic\_stage$
```{r,message=FALSE, warning=FALSE}

collapseStage = ifelse( clinical$ajcc_pathologic_stage %in% c("Stage I", "Stage IA", "Stage IB"),  "early", "late")

# ## alternatively
# collapseStage = ifelse( clinical$ajcc_pathologic_t %in%c("T1", "T1a", "T1b"),  "early", "late")

dat = data.frame(y = ifelse(clinical$Cluster == "1", 0, 1),
                 x = collapseStage)
res = glm(y~x, data = dat, family = "binomial")

print(summary(res))
confus = table(clinical$Cluster, collapseStage)
confus
OR = (19*20)/(7*10)
cat("Odds Ratio = ", OR, sep = "\t")
cat("Patients in Cluster 1 are mainly at early pathologic stage, while patients in Cluster 2 are mainly at late pathologic stage", sep = "\n")

```

